Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2018-07-07 survey/survey.tex
making a final pass

\begin{abstract}
  Our main models of computation (the Turing Machine and the RAM) and
  most modern computer architectures make fundamental assumptions
  about which primitive operations are realizable on a physical
  computing device. The consensus is that these primitive operations
  include logical operations like conjunction, disjunction and
  negation, as well as reading and writing to a large collection to
  memory locations. This perspective conforms to a macro-level view of
  physics and indeed these operations are realizable using macro-level
  devices involving thousands of electrons. This point of view is
  however incompatible with computation realized using quantum devices
  or analyzed using elementary thermodynamics as both these
  fundamental physical theories imply that information is a conserved
  quantity of physical processes and hence of primitive computational
  operations.

@@

No doubt, this consensus on the available primitive physical
operations has been successful. And no doubt, these operations
\emph{can} indeed be performed on a computer. Yet, today, with a
possible quantum computing revolution in sight and with unprecedented
explosion in embedded computers and cyber-physical systems, there are
reasons to re-think this foundational question again. In fact, the
calls to re-think this foundational question have been proclaimed by
physicists more than forty years ago as the following two quotes
testify:

@@

Bijections on \fin{n} have a specific name: permutations. As is
well-known, permutations can be generated by sequential compositions of
transpositions. Thus we can create a small language for writing
permutations on $\fin{n}$ as
\[\begin{array}{rcl}
p^n &::=& \mathit{id} ~\mid~ \mathit{swap}~i~j ~\mid~ p^n \fatsemi p^n
  \end{array}\]
where $i,j:\Nat$, $i\neq j$ and $i,j < n$. Note that we could remove
$\mathit{id}$ from the language and drop the $i\neq j$ condition so that
$\mathit{swap}~j~j$ would represent the identity permutation.

@@

In the figure (copied from a more general paper that includes
alternative designs~\cite{Rentergem2005OptimalDO}), the full adder
takes 4 inputs: the two bits to add $A$ and $B$, an incoming carry bit
$C_i$, and a heap input initialized to 0 to maintain
reversibility. There are four outputs: the first two are identical to
the incoming bits $A$ and $B$ and are considered ``garbage.'' The
third bit $S$ is the sum and the last bit $C_o$ is the outgoing carry
bit. In the notation used to describe the circuit, the $\times$ denotes
boolean negation, the dots are control bits. In our reversible
language, we can express this circuit as the following permutation of
type $[2^4] \rightarrow [2^4]$:

@@

\begin{prop}
The structure $\left(\Nat, \false, \true, \lor, \land\right)$
is a commutative semiring.
\end{prop}

We would like to adapt the commutative semiring definition to the
setting of structured types. First, types do not naturally want to be
put together into a ``set.''  This can be fixed if we replace the set
$R$ with a universe $U$, and replace the set membership $0 \in R$ with
the typing judgement $\bot : U$ (and similarly for the other
items). Our next instinct would be to similarly replace $=$ with a
type $A \equiv B$ that asserts that $A$ and $B$ are \emph{propositionally
equal}, i.e. reduce to equivalent type-denoting expressions under the rules of
the host type system.  This is however not true: the
proposition $A \times B \equiv B \times A$ is not
normally\footnote{Except in univalent type theory where equivalent
types are identified.} provable for arbitrary types $A$ and $B$. But
it should be clear that $A \times B$ and $B \times A$ contain
equivalent information. In other words, we would like to be able to
witness that $A \times B$ can be reversibly deformed into $B \times
A$, and vice-versa. Which motivates the introduction of type \emph{equivalences}. To
do this, we need a few important auxiliary concepts.

@@

In Martin-L\"{o}f Type Theory, normal forms mean $\beta$-reduced
$\eta$ long terms under $\alpha$-equivalence. In other words,
expressions are evaluated as much as possible ($\beta$-reduced),
all functions are fully applied ($\eta$ long), and the exact names
of bound variables are irrelevant ($\alpha$-equivalence). Note that
the above definition applies equally well to expressions that denote
values and expressions that denote types.

@@

\noindent
Of course, one does not get a programming language with just typed
constants! There is a need to perform multiple equivalences. There are
in fact three ways to do this: sequential composition $\odot$, choice
composition $\oplus$ (sometimes called juxtaposition), and parallel
composition $\otimes$. See Fig.~\ref{pi-combinators} for the
signatures. The construction $c_1 \odot c_2$ corresponds to performing
$c_1$ first, then $c_2$, and is the usual notion of composition -- and
corresponds to $\fatsemi$ of the language of permutations of
Sec.~\ref{sec:dataone}. The construction $c_1 \oplus c_2$ chooses to
perform $c_1$ or $c_2$ depending on whether the input is labelled
$\textsf{inl}$ or $\textsf{inr}$ respectively. Finally the
construction $c_1 \otimes c_2$ operates on a product structure, and
applies $c_1$ to the first component and $c_2$ to the second. The
language of permutations lacked the ability to combine permutations by
taking sums and products, which led to the awkward non-compositional
programming style illustrated in the full adder example~\ref{eq:adder}.

Thus the denotation of the $\Pi$ terms \emph{should} be permutations. But
given types $A$ and $B$ denoting $\fin{m}$ and $\fin{n}$ respectively,
what are $A \uplus B$ and $A \times B$ ?  They correspond exactly to
$\fin{m+n}$ and $\fin{m*n}$!  Geometrically, this corresponds to
concatenation for $A + B$, i.e. lining up the elements of $A$ first,
and then those of $B$. For $A * B$, one can picture this as
lining up the elements of $A$ horizontally, those of $B$ vertically
and perpendicular to those of $A$, and filling in the square with
pairs of elements from $A$ and $B$; if one re-numbers these sequentially,
reading row-wise, this gives an enumeration of $\fin{m*n}$.

@@

Since there are no values that have the type \ensuremath{0}, the reductions for
the combinators \identlp, \identrp, \identlsp, \identrsp omit the impossible
cases. \factorzr and \factorzl likewise do not appear as they have not
possible cases at all. However, \absorbr and \absorbl are treated slightly
differently: rather than \emph{eagerly} assuming they are impossible,
the purported inhabitant of $0$ given on one side is passed on to the other
side. The reason for this choice will have to wait for Sec.~\ref{langeqeq}
when we explain some higher-level symmetries (see Fig.~\ref{figc}).

@@

  \begin{itemize}
  \item For primitive isomorphisms \ensuremath{c}, \ensuremath{!~ c} is given by its
    inverse from Figures~\ref{pi-terms} and~\ref{more-pi}.

@@

\noindent We can further define that two combinators are 
\emph{observationally equivalent} if only all values of their
common domain, they evaluate to identical values.  More precisely,
we will say that for combinators $c_1, c_2 : b_1 \leftrightarrow b_2$,
$c_1~=~c_2$ whenever
\[
  \forall  ~v_1:b_1, v_2 : b_2. ~~ c_1 ~v_1 \mapsto v_2 \text{if and only if\ } c_2 ~v_1 \mapsto v_2
\]

Previously we had seen that types $b$ correspond to natural numbers,
more precisely~$|b|$, the size of $b$.  We had previously established that 
for any natural number $n$, there is a canonical set of size $n$,
which we denoted $[n]$. Furthermore, we can also define a
canonical \emph{type} that size, which we will denote $\sharp b$,
i.e. $\sharp b$ is a canonical type of size $|b|$.

\begin{definition}($\sharp$). By recursion on $|b|$.  First define a
type associated to each size. We will revert to using type notation
for greater clarity of this definition:
\[\begin{array}{rcl}
  \tau~ 0 & = & \bot \\
  \tau~ (1 + n) & = & \top \uplus \tau~ n \\
 \end{array}\]
\noindent so that we can define $\sharp b = \tau~ |b|$.
\end{definition}

We are now ready to go further and establish
that there is always an equivalence between.

\begin{proposition}
  For any type \ensuremath{b} there exists an isomorphism \ensuremath{b \leftrightarrow \sharp b}.
  \begin{proof}
    The fact that such an isomorphism exists is evident from the
    definition of size and what it means for two types to be
    isomorphic. While many equivalent constructions are possible
    for any type \ensuremath{b}, one such construction is given by 
    \sem{b}:

\[\begin{array}{rclr}
 \sem{0} & = & \idc \\
 \sem{1} & = & \idc \\
 \sem{1+b} & = & \idc \oplus \sem{b} \\
 \sem{(b_1+b_2)+b_3} & = & \assocrp \odot \sem{b_1 + (b_2 + b_3)} \\
 \sem{b_1 + b_2} & = & (\sem{b_1} \oplus \idc) \odot \sem{ \sharp b_1 + b_2 } \\
 \sem{0 * b_2} & = & \absorbr \\
 \sem{1 * b_2} & = & \identlt \odot \sem{b_2} \\
 \sem{(b_1 * b_2) * b_3} & = & \assocrt \odot \sem{b_1 * (b_2 * b_3)} \\
 \sem{(b_1+b_2) * b_3} & = & \dist \odot \sem{b_1 * b_3+b_2 * b_3} \\
 \end{array}\]
%originally subcode source isomorphisms.tex:598

@@

Combinators of \ensuremath{\Pi } can be written in terms of the
operators described previously or via a graphical language similar in
spirit to those developed for Geometry of Interaction
\cite{DBLP:conf/popl/Mackie95} and string diagrams for category
theory~\cite{BLUTE1996229,selinger-graphical}. Modulo some conventions
and shorthand we describe here, the wiring diagrams are equivalent to
the operator based (syntactic) description of programs.
\ensuremath{\Pi } combinators expressed in this graphical language
look like ``wiring diagrams''. Values take the form of ``particles''
that flow along the wires. Computation is expressed by the flow of
particles.

@@

\item
The product type \ensuremath{b_1 * b_2} may be represented both as one wire
labeled \ensuremath{b_1 * b_2} or by two parallel wires labeled \ensuremath{b_1} and
\ensuremath{b_2}. Both representations may be used interchangeably.

@@

There are two forms of parallel composition -- combinators
  can be combined additively \ensuremath{c_1+ c_2} (shown on the left) or
  multiplicatively \ensuremath{c_1 * c_2} (shown on the right).

@@

\noindent
\textit{Example.} As an example consider the wiring diagram
of the combinator \ensuremath{c} below:
\begin{flalign*}
& \begin{aligned}
c & : & b * (1+1) \leftrightarrow b + b \\
c & = & \swapt \odot \dist \odot (\identlt + \identlt) \\
\end{aligned} &&
\end{flalign*}

@@

Figure~\ref{type-isos} introduces our desired denotational
semantics, and section~\ref{sec:opsem} a direct definition
of an operational semantics. One obvious question arises:
do these correspond?

We can certainly associate to each $\Pi$ combinator an
equivalence between the denotation of each type%
\footnote{This is extracted from the Agda formalization
of this work, which has been reported on in~\cite{Carette2016}.}:
\begin{code}\hspace*{-4mm}
\>\AgdaFunction{c2equiv} \AgdaSymbol{:} \AgdaSymbol{\{}\AgdaBound{t₁}
\AgdaBound{t₂} \AgdaSymbol{:} \AgdaDatatype{U}\AgdaSymbol{\}} \AgdaSymbol{→}
\AgdaSymbol{(}\AgdaBound{c} \AgdaSymbol{:} \AgdaBound{t₁} \AgdaDatatype{⟷}
\AgdaBound{t₂}\AgdaSymbol{)} \AgdaSymbol{→} \AgdaFunction{⟦} \AgdaBound{t₁}
\AgdaFunction{⟧} \AgdaFunction{≃} \AgdaFunction{⟦} \AgdaBound{t₂}
\AgdaFunction{⟧}%
\end{code}

@@

We will however show that one can program all logical gates in
$\Pi$. We will start with a few simply examples and then discuss the
expressiveness of the language and discuss various properties.

@@

\ensuremath{\mathit{reverse} = \swapt \odot (\swapt  *  \idc)~ \odot \assocrt}

\noindent We can check that \ensuremath{\mathit{reverse}} does the right thing by
applying it to a value \ensuremath{(v_1, (v_2, v_3))} and writing out the full
derivation tree of the reduction.  The combinator \ensuremath{\mathit{reverse}}, like
many others we will see in this thesis, is formed by sequentially
composing several simpler combinators. Instead of presenting the
operation of \ensuremath{\mathit{reverse}} as a derivation tree, it is easier (purely
for presentation reasons) to flatten the tree into a sequence of
reductions as caused by each component. Such a sequence of reductions
is given below:
\[\begin{array}{rlr}
 & (v_1, (v_2, v_3)) \\
 \swapt & ((v_2, v_3), v_1) \\
 \swapt  *  \idc & ((v_3, v_2), v_1) \\
 \assocrt & (v_3, (v_2, v_1)) \\
 \end{array}\]
%subcode source isomorphisms.tex:979

@@

\ensuremath{\mathit{if}_{c_1,c_2} = \dist \odot ((\idc  *  c_1) + (\idc  *  c_2)) \odot \factor}

>>>>

\begin{abstract}
  Our main models of computation (the Turing Machine and the RAM) and
  most modern computer architectures make fundamental assumptions
  about which primitive operations are realizable on a physical
  computing device. The consensus is that these primitive operations
  include logical operations like conjunction, disjunction and
  negation, as well as reading and writing to a large collection of
  memory locations. This perspective conforms to a macro-level view of
  physics and indeed these operations are realizable using macro-level
  devices involving thousands of electrons. This point of view is
  however incompatible with computation realized using quantum devices
  or analyzed using elementary thermodynamics as both these
  fundamental physical theories imply that information is a conserved
  quantity of physical processes and hence of primitive computational
  operations.

@@

No doubt, this consensus on the available primitive physical
operations has been successful. And no doubt, these operations
\emph{can} indeed be performed on a computer. Yet, today, with a
possible quantum computing revolution in sight and with unprecedented
explosion in embedded computers and cyber-physical systems, there are
reasons to re-think this foundational question again. In fact, the
calls to re-think this foundational question have been proclaimed by
physicists almost forty years ago as the following two quotes testify:

@@

Bijections on \fin{n} have a specific name: permutations. As is
well-known, permutations can be generated by sequential compositions of
transpositions. Thus we can create a small language for writing
permutations on $\fin{n}$ as:
\[\begin{array}{rcl}
p^n &::=& \mathit{id} ~\mid~ \mathit{swap}~i~j ~\mid~ p^n \fatsemi p^n
  \end{array}\]
where $i,j:\Nat$, $i\neq j$ and $i,j < n$. Note that we could remove
$\mathit{id}$ from the language and drop the $i\neq j$ condition so that
$\mathit{swap}~j~j$ would represent the identity permutation.

@@

In the figure (copied from a more general paper that includes
alternative designs~\cite{Rentergem2005OptimalDO}), the full adder
takes 4 inputs: the two bits to add $A$ and $B$, an incoming carry bit
$C_i$, and a heap input initialized to 0 to maintain
reversibility. There are four outputs: the first two are identical to
the incoming bits $A$ and $B$ and are considered ``garbage.'' The
third bit $S$ is the sum and the last bit $C_o$ is the outgoing carry
bit. In the notation used to describe the circuit, the $\times$ denotes
boolean negation and the dots are control bits. In our reversible
language, we can express this circuit as the following permutation of
type $[2^4] \rightarrow [2^4]$:

@@

\begin{prop}
The structure $\left(\{\false,\true\}, \false, \true, \lor, \land\right)$
is a commutative semiring.
\end{prop}

We would like to adapt the commutative semiring definition to the
setting of structured types. First, types do not naturally want to be
put together into a ``set.''  This can be fixed if we replace the set
$R$ with a universe $U$, and replace the set membership $0 \in R$ with
the typing judgement $\bot : U$ (and similarly for the other
items). Our next instinct would be to similarly replace $=$ with a
type $A \equiv B$ that asserts that $A$ and $B$ are
\emph{propositionally equal}, i.e. reduce to equivalent type-denoting
expressions under the rules of the host type system.  This is however
not true: the proposition $A \times B \equiv B \times A$ is not
normally\footnote{Except in univalent type theory where equivalent
  types are identified.} provable for arbitrary types $A$ and $B$. But
it should be clear that $A \times B$ and $B \times A$ contain
equivalent information. In other words, we would like to be able to
witness that $A \times B$ can be reversibly deformed into
$B \times A$, and vice-versa, which motivates the introduction of type
\emph{equivalences}. To do this, we need a few important auxiliary
concepts.

@@

In Martin-L\"{o}f Type Theory, normal forms mean $\beta\eta$-long
normal forms under $\alpha$-equivalence. In other words, expressions
are evaluated as much as possible ($\beta$-reduced), all functions are
fully applied ($\eta$-long), and the exact names of bound variables
are irrelevant ($\alpha$-equivalence). Note that the above definition
applies equally well to expressions that denote values and expressions
that denote types.

@@

\noindent
Of course, one does not get a programming language with just typed
constants! There is a need to perform multiple equivalences. There are
in fact three ways to do this: sequential composition $\odot$, choice
composition $\oplus$ (sometimes called juxtaposition), and parallel
composition $\otimes$. See Fig.~\ref{pi-combinators} for the
signatures. The construction $c_1 \odot c_2$ corresponds to performing
$c_1$ first, then $c_2$, and is the usual notion of composition -- and
corresponds to $\fatsemi$ of the language of permutations of
Sec.~\ref{sec:dataone}. The construction $c_1 \oplus c_2$ chooses to
perform $c_1$ or $c_2$ depending on whether the input is labelled
$\textsf{inl}$ or $\textsf{inr}$ respectively. Finally the
construction $c_1 \otimes c_2$ operates on a product structure, and
applies $c_1$ to the first component and $c_2$ to the second. The
language of permutations lacked the ability to combine permutations by
taking sums and products, which led to the awkward non-compositional
programming style illustrated in the full adder
example~(Eq.~\ref{eq:adder}).

Thus the denotation of the $\Pi$ terms \emph{should} be permutations. But
given types $A$ and $B$ denoting $\fin{m}$ and~$\fin{n}$ respectively,
what are $A \uplus B$ and $A \times B$ ?  They correspond exactly to
$\fin{m+n}$ and $\fin{m*n}$!  Geometrically, this corresponds to
concatenation for $A + B$, i.e. lining up the elements of $A$ first,
and then those of~$B$. For $A * B$, one can picture this as
lining up the elements of $A$ horizontally, those of $B$ vertically
and perpendicular to those of $A$, and filling in the square with
pairs of elements from $A$ and $B$; if one re-numbers these sequentially,
reading row-wise, this gives an enumeration of $\fin{m*n}$.

@@

Since there are no values that have the type \ensuremath{0}, the
reductions for the combinators \identlp, \identrp, \identlsp, and
\identrsp\ omit the impossible cases. \factorzr\ and \factorzl\
likewise do not appear as they have no possible cases at all. However,
\absorbr\ and \absorbl\ are treated slightly differently: rather than
\emph{eagerly} assuming they are impossible, the purported inhabitant
of $0$ given on one side is passed on to the other side. The reason
for this choice will have to wait for Sec.~\ref{langeqeq} when we
explain some higher-level symmetries (see Fig.~\ref{figc}).

@@

  \begin{itemize}
  \item For primitive isomorphisms \ensuremath{c}, \ensuremath{!~ c} is given by its
    inverse from Figs.~\ref{pi-terms} and~\ref{more-pi}.

@@

\noindent We can further define that two combinators are
\emph{observationally equivalent} if on all values of their common
domain, they evaluate to identical values.  More precisely, we will
say that for combinators $c_1, c_2 : b_1 \leftrightarrow b_2$,
$c_1~=~c_2$ whenever:
\[
  \forall  ~v_1:b_1, v_2 : b_2. ~~ c_1 ~v_1 \mapsto v_2 ~\text{if and only if\ } c_2 ~v_1 \mapsto v_2
\]

Each type $b$ has a size $|b|$ defined in the obvious way. We had
previously established that for any natural number $n$, there is a
canonical set of size $n$, which we denoted $[n]$. Furthermore, we can
also define a canonical \emph{type} of that size, which we will denote
$\sharp\, b$, i.e. $\sharp\, b$ is a canonical type of size $|b|$.

\begin{definition}($\sharp$). By recursion on $|b|$.  First define a
type associated to each size. We will revert to using type notation
for greater clarity of this definition:
\[\begin{array}{rcl}
  \tau~ (0) & = & \bot \\
  \tau~ (1 + n) & = & \top \uplus \tau~ (n) \\
 \end{array}\]
\noindent so that we can define $\sharp\, b = \tau~ |b|$.
\end{definition}

We are now ready to go further and establish
that there is always an equivalence between a type and the canonical
type of the same size.

\begin{proposition}
  For any type \ensuremath{b} there exists an isomorphism \ensuremath{b \leftrightarrow \sharp\, b}.
  \begin{proof}
    The fact that such an isomorphism exists is evident from the
    definition of size and what it means for two types to be
    isomorphic. While many equivalent constructions are possible
    for any type \ensuremath{b}, one such construction is given by 
    \sem{b}:

\[\begin{array}{rclr}
 \sem{0} & = & \idc \\
 \sem{1} & = & \idc \\
 \sem{1+b} & = & \idc \oplus\ \sem{b} \\
 \sem{(b_1+b_2)+b_3} & = & \assocrp \odot \sem{b_1 + (b_2 + b_3)} \\
 \sem{b_1 + b_2} & = & (\sem{b_1} \oplus \idc) \odot \sem{ \sharp\, b_1 + b_2 } \\
 \sem{0 * b_2} & = & \absorbr \\
 \sem{1 * b_2} & = & \identlt \odot \sem{b_2} \\
 \sem{(b_1 * b_2) * b_3} & = & \assocrt \odot \sem{b_1 * (b_2 * b_3)} \\
 \sem{(b_1+b_2) * b_3} & = & \dist \odot \sem{b_1 * b_3+b_2 * b_3} \\
 \end{array}\]
%originally subcode source isomorphisms.tex:598

@@

Combinators of \ensuremath{\Pi } can be written in terms of the
operators described previously or via a graphical language similar in
spirit to those developed for Geometry of Interaction
\cite{DBLP:conf/popl/Mackie95} and string diagrams for category
theory~\cite{BLUTE1996229,selinger-graphical}. Modulo some conventions
and shorthand we describe here, the wiring diagrams are equivalent to
the operator based (syntactic) description of programs.
\ensuremath{\Pi } combinators expressed in this graphical language
look like ``wiring diagrams.'' Values take the form of ``particles''
that flow along the wires. Computation is expressed by the flow of
particles.

@@

\item
The product type \ensuremath{b_1 * b_2} may be represented both as one wire
labeled \ensuremath{b_1 \times b_2} or by two parallel wires labeled \ensuremath{b_1} and
\ensuremath{b_2}. Both representations may be used interchangeably.

@@

There are two forms of parallel composition -- combinators
  can be combined additively \ensuremath{c_1 \oplus\ c_2} (shown on the left) or
  multiplicatively \ensuremath{c_1 \otimes c_2} (shown on the right).

@@

\noindent
\textit{Example.} As an example consider the wiring diagram
of the combinator \ensuremath{c} below:
\[\begin{array}{rcl}
c & : & b * (1+1) \leftrightarrow b + b \\
c & = & \swapt \odot \dist \odot (\identlt \oplus \identlt) 
\end{array}\]

@@

Fig.~\ref{type-isos} introduces our desired denotational semantics,
and Sec.~\ref{sec:opsem} is a direct definition of an operational
semantics. One obvious question arises: do these correspond?

We can certainly associate to each $\Pi$ combinator an
equivalence between the denotation of each type%
\footnote{This is extracted from the Agda formalization
of this work, which has been reported on in a previous paper~\cite{Carette2016}.}:
\begin{code}\hspace*{-4mm}
\>\AgdaFunction{c2equiv} \AgdaSymbol{:} \AgdaSymbol{\{}\AgdaBound{t₁}
\AgdaBound{t₂} \AgdaSymbol{:} \AgdaDatatype{U}\AgdaSymbol{\}} \AgdaSymbol{→}
\AgdaSymbol{(}\AgdaBound{c} \AgdaSymbol{:} \AgdaBound{t₁} \AgdaDatatype{⟷}
\AgdaBound{t₂}\AgdaSymbol{)} \AgdaSymbol{→} \AgdaFunction{⟦} \AgdaBound{t₁}
\AgdaFunction{⟧} \AgdaFunction{≃} \AgdaFunction{⟦} \AgdaBound{t₂}
\AgdaFunction{⟧}%
\end{code}

@@

We will however show that one can program all logical gates in
$\Pi$. We will start with a few simple examples and then discuss the
expressiveness of the language and its properties.

@@

\ensuremath{\mathit{reverse} = \swapt \odot (\swapt  \otimes  \idc)~ \odot \assocrt}

\noindent We can check that \ensuremath{\mathit{reverse}} does the right thing by
applying it to a value \ensuremath{(v_1, (v_2, v_3))} and writing out the full
derivation tree of the reduction.  The combinator \ensuremath{\mathit{reverse}}, like
many others we will see in this paper, is formed by sequentially
composing several simpler combinators. Instead of presenting the
operation of \ensuremath{\mathit{reverse}} as a derivation tree, it is easier (purely
for presentation reasons) to flatten the tree into a sequence of
reductions as caused by each component. Such a sequence of reductions
is given below:
\[\begin{array}{rlr}
 & (v_1, (v_2, v_3)) \\
 \swapt & ((v_2, v_3), v_1) \\
 \swapt \otimes  \idc & ((v_3, v_2), v_1) \\
 \assocrt & (v_3, (v_2, v_1)) \\
 \end{array}\]
%subcode source isomorphisms.tex:979

@@

\ensuremath{\mathit{if}_{c_1,c_2} = \dist \odot ((\idc  \otimes\  c_1) + (\idc \otimes\  c_2)) \odot \factor}
